---
title: "STAT/MATH 495: Problem Set 09"
author: "Sarah Teichman"
date: "2017-11-07"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    collapsed: false
    smooth_scroll: false
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.width=8, fig.height=4.5, message=FALSE, warning = FALSE
  )
set.seed(50)

library(tidyverse)
library(proxy)
```

# Collaboration

Please indicate who you collaborated with on this assignment: 

# Question 1: Run k-means

Loading in data
```{r}
observations_1 <- read_csv("data/observations_1.csv")
observations_2 <- read_csv("data/observations_2.csv")
```

Exploratory Data Analysis
```{r}
ggplot(observations_1, aes(x=x1,y=x2)) +
  geom_point() +
  labs(title="Distribution from observations_1")
ggplot(observations_2, aes(x=x1,y=x2)) +
  geom_point() + 
  labs(title="Distribution from observations_2")
```

Creating function to run K-means clustering (with observations as the argument)
```{r}
# Fit model for k=2
kmc <- function(observations) {
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)

return(cluster_centers)
}
```

Running the function for observations 1 and observations 2
```{r}
cluster_centers_1 <- data.frame(numeric(20),numeric(20),numeric(20))
cluster_centers_2 <- data.frame(numeric(20),numeric(20),numeric(20))
names(cluster_centers_1) <- c('x1','x2','cluster')
names(cluster_centers_2) <- c('x1','x2','cluster')
for (i in 1:10) {
  x <- kmc(observations_1) 
  cluster_centers_1$x1[i] <- x$x1[1]
  cluster_centers_1$x2[i] <- x$x2[1]
  cluster_centers_1$cluster[i] <- 1
  cluster_centers_1$x1[10+i] <- x$x1[2]
  cluster_centers_1$x2[10+i] <- x$x2[2]
  cluster_centers_1$cluster[10+i] <- 2
}
cluster_centers_1 %>%
  group_by(x1,x2) %>%
  summarize(count=n())
for (i in 1:10) {
  x <- kmc(observations_2) 
  cluster_centers_2$x1[i] <- x$x1[1]
  cluster_centers_2$x2[i] <- x$x2[1]
  cluster_centers_2$cluster[i] <- 1
  cluster_centers_2$x1[10+i] <- x$x1[2]
  cluster_centers_2$x2[10+i] <- x$x2[2]
  cluster_centers_2$cluster[10+i] <- 2
}
cluster_centers_2 %>%
  group_by(x1,x2) %>%
  summarize(count=n())
```

**Questions**:

1. Run KMC 10 times on `observations_1` and comment on the consistency of the
results.
2. Speculate on the root cause of any consistency or inconsistency in the
results.
3. Run KMC 10 times on `observations_2` and comment on the consistentcy of the
results.
4. Speculate on the root cause of any consistency or inconsistency in the
results.

**Answers**:

1. The results are not very consistent (over the 10 trials, there are about 4-6 different points that are identified as cluster centers).
2. The results aren't consistent because the points are spread out throughout the two attributes, and don't appear to be in any meaningful clusters.
3. The results are very consistent (over the 10 trials, the same two cluster centers are identified every time).
4. The results are consistent because the points appear to be split in two clear groups, one that has smaller values of both x1 and x2, and one that has larger values for both of these attributes. 

# Bonus question: Code your own

Read ISLR page 388 Algorithm 10.1 and implement k-means clustering from scratch.
Don't worry about doing it for general $k$; keep it simple and do it for $k=2$
specifically. Apply it to `observations_2` from above.

```{r}
get_cluster <- function(A,B) {
  distance_matrix <- proxy::dist(x=A, y=B)
  clusters <- apply(distance_matrix, 1, which.min)
  return(clusters)
}
```

```{r}
all_equal <- function(A,B,n) {
  for (i in 1:n) {
    if (A[i] != B[i]) {
      return(FALSE)
    }
  }
  return(TRUE)
}
```

```{r}
num <- data.frame(seq(1,100))
x <- sample_n(num,50)
#setting intial random clusters
observations <- observations_2 %>%
  mutate(id = seq(1:100),
         cluster = ifelse(id %in% x$seq.1..100., 1,2))
#finding cluster centers
observations <- observations %>%
  mutate(cluster_old = numeric(nrow(observations)),
         cluster_new = cluster)
counter = 0
while(all_equal(observations$cluster_old,observations$cluster_new,nrow(observations))==FALSE) {
  counter = counter + 1
  info <- observations %>%
    group_by(cluster_new) %>%
    summarise(x1 = mean(x1),x2 = mean(x2))
  obs_dist <- observations %>%
    select(x1,x2)
  cent_dist <- info %>%
    select(x1,x2)
  observations$cluster_old = observations$cluster_new
  clusters <- get_cluster(obs_dist,cent_dist)
  observations$cluster_new <- clusters 
}
observations <- observations %>%
  mutate(cluster = cluster_new) %>%
  select(id,cluster)
head(observations)
tail(observations)
counter
```

